### 2023-06-20 bachelorthesis/03-abstract.tex
#39: Bachelor thesis finished

Ion temperature gradient driven turbulence (ITG) close to marginal stability exhibits zonal flow pattern formation on mesoscales, so-called $\exb$ staircase structures. Such pattern formation has been observed in local gradient-driven flux-tube simulations as well as global gradient-driven and global flux-driven studies. \bigskip

@@

Furthermore, it is shown by multiple box size convergence scans that a mesoscale pattern  size of $\sim 57-76\,\rhoth$ is inherent to ITG driven turbulence with Cyclone Base Case parameters in the local limit. This outcome also implies that a typical scale for avalanche-like transport is inherent to ITG driven turbulence.

>>>>

Ion temperature gradient-driven turbulence (ITG) close to marginal stability exhibits zonal flow pattern formation on mesoscales, so-called $\exb$ staircase structures. Such pattern formation has been observed in local gradient-driven flux-tube simulations as well as global gradient-driven and global flux-driven studies. \bigskip

@@

Furthermore, it is shown by multiple box size convergence scans that a mesoscale pattern  size of $\sim 57-76\,\rhoth$ is inherent to ITG-driven turbulence with Cyclone Base Case parameters in the local limit. This outcome also implies that a typical scale for avalanche-like transport is inherent to ITG-driven turbulence.

### 2023-06-20 bachelorthesis/21-motion.tex
#39: Bachelor thesis finished

Due to the Lorentz force, particles with a velocity component perpendicular to the homogenous magnetic field $v_{\perp}$ undergo a circular motion in the plane perpendicular to the magnetic field. This type of motion has circular frequency, which is often referred to as \textit{cyclotron frequency} and is defined as
\begin{gather}
    \omega_\mathrm{c} = \frac{|q|B}{m}~,
\end{gather}
where $m$ and $q$ are the mass and the charge of the particle and $B$ the strength of the magnetic field. The radius, the so called \textit{Larmor radius}, of this motion is given by
\begin{gather}
    \rho_\mathrm{L} = \frac{mv_{\perp}}{|q|B}
\end{gather}
with the center often being referred to as \textit{gyrocenter}. Note that since the Lorentz force depends on the species charge of the particle, the circulation direction is the opposite between electron in ions.\\\bigskip

@@

When an electric field with a component parallel to the magnetic field $E_\parallel$ influences the plasma the charged particles are accelerated by the electric force
\begin{gather}
    F_{\parallel,E} = qE_\parallel~.
\end{gather}
The parallel motion follows then from the equation of motion. Here the direction of the motion also depends on the species type.\\\bigskip

Since magnetic fields are not always homogenous, an inhomogeneous magnetic field with its gradient $\nabla B$ containing a component parallel to the magnetic field which is given by
\begin{gather}
    \nabla_{\!\parallel} B = \frac{\vect{B}}{B} \cdot \nabla B 
\end{gather}
causes the force
\begin{gather}
    F_{\parallel,\nabla_{\!\parallel} B} = - \frac{mv^2_{\perp}}{2B} \nabla_{\!\parallel} B = \mu \nabla_{\!\parallel} B~;~~~~\mu = \frac{mv^2_{\perp}}{2B}
\end{gather}
with \textit{magnetic moment} $\mu$. The magnetic moment $\mu$ is an adabatic invariant (constant of motion) if the variation of the magnetic field over time is smaller than the inverse of the cyclotron freqency $\omega^{-1}_\mathrm{c}$ and the spatial variation is larger the Larmor radius $\rho_\mathrm{L}$. The resulting force has its application in the mirror effect where a charged particle gets reflected due to this force. \cite{Wesson2011}

@@

In the presence of a magnetic field (homogenous, inhomogeneous or perturbed) and electric fields the gyrocenter undergoes slow (compared to the thermal velocity $\vth$) drift motions perpendicular to the magnetic field. There are serval examples for this drift motion. According to this thesis topic only the main three drift types will be covered in the following.

\begin{enumerate}
    \item {\boldmath $\exb$} \textbf{Drift:}\\
    If an electric field $\vect{E}$ with a perpendicular component together with the magnetic field $\vect{B}$ (both fields are homogenous) is present the acting Coulomb force and Lorentz force results into a drift of the gyrocenter with
    \begin{gather}
        \vect{v}_{E} = \frac{\vect{E}\times\vect{B}}{B^2}
        \label{eq:exb}
    \end{gather}
    which is called the $\exb$ drift. Since both acting forces direction depends on the species type the direction of the $\exb$ drift is for every species the same.
    \item {\boldmath $\nabla B$} \textbf{Drift:}\\
    Inhomogeneous magnetic field causes a gradient $\nabla B$ of the magnetic field. Because of that gradient the gyrocenter undergoes a $\nabla B$ drift defined by
    \begin{gather}
        \vect{v}_{\nabla B} = \frac{m v^2_{\perp}}{2 q}\frac{\vect{B}\times \nabla B}{B^3}~.
        \label{eq:gradB}
    \end{gather}
    The gradient of the magnetic field $\nabla B$ varies thereby on scales larger compared to the Larmor radius. The direction of the $\nabla B$ drift depends on the species type.
    \item \textbf{Curvature Drift:}\\
    Due to centrifugal force acting on the particle in a curved magnetic field the gyrocenter experiences a curvature drift according to
    \begin{gather}
        \vect{v}_{\kappa} = \frac{m v^2_{\parallel}}{q}\frac{\vect{B}\times \vect{\kappa}}{B^2} = \frac{m v^2_{\parallel}}{q} \frac{\vect{B}\times \nabla B}{B^3}~;\qquad\vect{\kappa} = -(\vect{b}\cdot \nabla)\vect{b} = \frac{\nabla B}{B}~,
        \label{eq:curvature}
    \end{gather}
    where $\vect{b}$ is the unit vector along the magnetic field. To obtain the result for the curvature $\vect{\kappa}$ in Eq. (\ref{eq:curvature}) the plasma pressure has to be small compared to the magnetic field strength $B$. In the form of Eq. (\ref{eq:curvature}) $\nabla B$ and curvature drift can be treated similarly. \cite{Wesson2011}
\end{enumerate}

>>>>

Due to the Lorentz force, particles with a velocity component perpendicular to the homogenous magnetic field $v_{\perp}$ undergo a circular motion in the plane perpendicular to the magnetic field [Fig. \ref{fig:perp-par-motion}(a)]. This type of motion has circular frequency, which is often referred to as \textit{cyclotron frequency} and is defined as
\begin{gather}
    \omega_\mathrm{c} = \frac{|q|B}{m}~,
\end{gather}
where $m$ and $q$ are the mass and the charge of the particle and $B$ the strength of the magnetic field. The radius, the so called \textit{Larmor radius}, of this motion is given by
\begin{gather}
    \rho_\mathrm{L} = \frac{mv_{\perp}}{|q|B}
\end{gather}
with the center often being referred to as \textit{gyrocenter}. Note that since the Lorentz force depends on the species charge of the particle, the circulation direction is the opposite between electron in ions.\\\bigskip

@@

When an electric field with a component parallel to the magnetic field $E_\parallel$ influences the plasma the charged particles are accelerated by the electric force
\begin{gather}
    F_{\parallel,E} = qE_\parallel~.
\end{gather}
The parallel motion follows then from the equation of motion. Here the direction of the motion also depends on the species type [Fig. \ref{fig:perp-par-motion}(b)].\\\bigskip

Since magnetic fields are not always homogenous, an inhomogeneous magnetic field with its gradient $\nabla B$ containing a component parallel to the magnetic field which is given by
\begin{gather}
    \nabla_{\!\parallel} B = \frac{\vect{B}}{B} \cdot \nabla B 
\end{gather}
causes the force
\begin{gather}
    F_{\parallel,\nabla_{\!\parallel} B} = - \frac{mv^2_{\perp}}{2B} \nabla_{\!\parallel} B = - \mu \nabla_{\!\parallel} B~;~~~~\mu = \frac{mv^2_{\perp}}{2B}
\end{gather}
with \textit{magnetic moment} $\mu$. The magnetic moment $\mu$ is an adabatic invariant (constant of motion) if the variation of the magnetic field over time is smaller than the inverse of the cyclotron frequency $\omega^{-1}_\mathrm{c}$ and the spatial variation is larger the Larmor radius $\rho_\mathrm{L}$. The resulting force has its application in the mirror effect where a charged particle gets reflected due to this force [Fig. \ref{fig:perp-par-motion}(c)]. \cite{Wesson2011}

\begin{center}
    \captionsetup{type=figure}
    \begin{tabular}{c c c c c c}
        \input{../pictures/Theory/Lorentz-Force_Thesis.tex} &
        \input{../pictures/Theory/Electric-Force.tex} &
        \input{../pictures/Theory/Inhomogenous-Magnetic-Field.tex}\\[0.3cm]
        \textbf{(a)} & \textbf{(b)} & \textbf{(c)} \\
    \end{tabular}
    \captionof{figure}{Forces acting on a charged particle:\\
        \begin{tabular}{l l}
            \textbf{(a)} & Lorentz force $\vect{F}_{\perp, B}$ perpendicular to velocity $\vect{v}$ which causes,\\
                         & circular motion with Lamor radius $\rho_\mathrm{L}$ and cyclotron frequency $\omega_\mathrm{c}$ \\
            \textbf{(b)} & Electric force $\vect{F}_{\parallel,E}^\pm$,\\
            \textbf{(c)} & Mirror effect with force $\vect{F}_{\parallel,\nabla_{\!\parallel} B}$ and magnetic moment $\mu$ \\
                         & caused by an inhomogeneous magnetic field $\vect{B}$.  \\
        \end{tabular}
    }
    \label{fig:perp-par-motion}
\end{center}

@@

In the presence of a magnetic field (homogenous, inhomogeneous or perturbed) and electric fields the gyrocenter undergoes slow (compared to the thermal velocity $\vth$) drift motions perpendicular to the magnetic field. There are several examples for this drift motion. According to this thesis topic only the main three drift types will be covered in the following.

\begin{enumerate}
    \item {\boldmath $\exb$} \textbf{Drift:}\\
    If an electric field $\vect{E}$ with a perpendicular component together with the magnetic field $\vect{B}$ (both fields are homogenous) is present the acting Coulomb force and Lorentz force results into a drift of the gyrocenter with
    \begin{gather}
        \vect{v}_{E} = \frac{\vect{E}\times\vect{B}}{B^2}
        \label{eq:exb}
    \end{gather}
    which is called the $\exb$ drift. Since both acting forces direction depends on the species type the direction of the $\exb$ drift is for every species the same [Fig. \ref{fig:drift}(a)].
    \item {\boldmath $\nabla B$} \textbf{Drift:}\\
    Inhomogeneous magnetic field causes a gradient $\nabla B$ of the magnetic field. Because of that gradient the gyrocenter undergoes a $\nabla B$ drift defined by
    \begin{gather}
        \vect{v}_{\nabla B} = \frac{m v^2_{\perp}}{2 q}\frac{\vect{B}\times \nabla B}{B^3}~.
        \label{eq:gradB}
    \end{gather}
    The gradient of the magnetic field $\nabla B$ varies thereby on scales larger compared to the Larmor radius. The direction of the $\nabla B$ drift depends on the species type [Fig. \ref{fig:drift}(b)].
    \item \textbf{Curvature Drift:}\\
    Due to centrifugal force acting on the particle in a curved magnetic field the gyrocenter experiences a curvature drift according to
    \begin{gather}
        \vect{v}_{\kappa} = \frac{m v^2_{\parallel}}{q}\frac{\vect{B}\times \vect{\kappa}}{B^2} = \frac{m v^2_{\parallel}}{q} \frac{\vect{B}\times \nabla B}{B^3}~;\qquad\vect{\kappa} = -(\vect{b}\cdot \nabla)\vect{b} = \frac{\nabla B}{B}~,
        \label{eq:curvature}
    \end{gather}
    where $\vect{b}$ is the unit vector along the magnetic field. To obtain the result for the curvature $\vect{\kappa}$ in Eq. (\ref{eq:curvature}) the plasma pressure has to be small compared to the magnetic field strength $B$. In the form of Eq. (\ref{eq:curvature}) $\nabla B$ and curvature drift can be treated similarly. \cite{Wesson2011}
\end{enumerate}

\begin{center}
    \captionsetup{type=figure}
    \begin{tabular}{c c}
        \input{../pictures/Theory/ExB-Drift.tex} &
        \input{../pictures/Theory/GradB-Drift.tex} \\[0.3cm]
        \textbf{(a)} & \textbf{(b)} \\
    \end{tabular}
    \captionof{figure}{Dirft motion in gyrocenter:\\
        \begin{tabular}{l l}
            \textbf{(a)} & $\exb$ Drift with drift velocity $\vect{v}_E$, electric field $\vect{E}$ and\\
                         & magnetic field $\vect{B}$,\\
            \textbf{(b)} & $\nabla B$ Drift/Curvature Drift with drift velocity $\vect{v}^{\pm}_{\nabla B;\kappa}$, magnetic\\
                         & field $\vect{B}$ and gradient of the magnetic field $\nabla B$. \\
        \end{tabular}
    }
    \label{fig:drift}
\end{center}

### 2023-06-20 bachelorthesis/23-iontempgrad.tex
#39: Bachelor thesis finished

\includegraphicsHere{Theory/ITG-Turbulence.pdf}{
	Sketch of ion temperature gradient driven instability on the outboard low field side of the tokamak \cite{Casson_PHD}
}{fig:ITG}{0.5}

The ITG driven instability is present when a sufficiently large ion temperature gradient occurs and is driven by the so-called bad curvature of the outboard low field side of the tokamak. The mechanism is briefly explained with the help of Fig. \ref{fig:ITG} and Ref. \citenum{Beer_PHD, Casson_PHD, Dannert_PHD}. An ion temperature perturbation ($T^\pm \gtrless 0$ \textcolor{blue}{blue} and \textcolor{red}{red} cells) on the outboard low field side of the tokamak causes the modulation of curvature drift and $\nabla B$ drift with velocity $\vect{v}_\mathrm{d}$. The divergence of the velocity $\vect{v}_\mathrm{d}$ is therefore not zero which results in compression of the plasma ($n_\pm \gtrless 0$, \textcolor{gray}{gray} cells). This relates to the perturbation of the potential ($\phi^\pm  \gtrless 0$) in the adiabatic electron response. As stated in the equation for the electric field ($\vect{E} = -\nabla \phi$) the perturbation of the potential $\phi$ causes an electric field $\vect{E}$ which results in a $\exb$ drift with velocity $\vect{v}_E$. The $\exb$ drift direction transports hot plasma in the outer hot region of perturbation ($T^+ > 0$, \textcolor{red}{red} cell) and cooler plasma into inner cold region perturbation ($T^- < 0$, \textcolor{blue}{blue} cell) which reinforce the process and instability grows.\bigskip

The described mechanism causes perturbation on the outboard low field side of the tokamak this region has a \enquote{bad curvature} and instability happens there. On the inboard high field side the opposing $\nabla B$ and $\nabla T$ results in a decay of perturbation which is referred to as \enquote{good curvature}. Furthermore, such turbulent structures are called \textit{eddies} which have a typical radial scale of several Larmor radii. \cite{Newins2006}
The initial perturbation is growing if the normalized temperature gradient overcomes a critical value $R/L_\mathrm{T,c}$. For nonlinear turbulence with Cyclone Base Case parameters and circular concentric flux-surfaces the critical value is given by $R/L_\mathrm{T,c} \sim 6$. \cite{Dimits2000, Isliker2010}

>>>>

\begin{center}
	\captionsetup{type=figure}
	\input{../pictures/Theory/ITG-Turbulence.tex}
	\captionof{figure}{Sketch of ion temperature gradient driven instability on the outboard low field side of the tokamak. \cite{Casson_PHD}}
	\label{fig:ITG}
\end{center}

The ITG-driven instability is present when a sufficiently large ion temperature gradient occurs and is driven by the so-called bad curvature of the outboard low field side of the tokamak. The mechanism is briefly explained with the help of Fig. \ref{fig:ITG} and Ref. \citenum{Beer_PHD, Casson_PHD, Dannert_PHD}. An ion temperature perturbation ($T^\pm \gtrless 0$ \textcolor{blue}{blue} and \textcolor{red}{red} cells) on the outboard low field side of the tokamak causes the modulation of curvature drift and $\nabla B$ drift with velocity $\vect{v}_\mathrm{d}$. The divergence of the velocity $\vect{v}_\mathrm{d}$ is therefore not zero which results in compression of the plasma ($n_\pm \gtrless 0$, \textcolor{gray}{gray} cells). This relates to the perturbation of the potential ($\phi^\pm  \gtrless 0$) in the adiabatic electron response. As stated in the equation for the electric field ($\vect{E} = -\nabla \phi$) the perturbation of the potential $\phi$ causes an electric field $\vect{E}$ which results in a $\exb$ drift with velocity $\vect{v}_E$. The $\exb$ drift direction transports hot plasma in the outer hot region of perturbation ($T^+ > 0$, \textcolor{red}{red} cell) and cooler plasma into inner cold region perturbation ($T^- < 0$, \textcolor{blue}{blue} cell) which reinforce the process and instability grows.
\newpage
The described mechanism causes perturbation on the outboard low field side of the tokamak this region has a \enquote{bad curvature} and instability happens there. On the inboard high field side the opposing $\nabla B$ and $\nabla T$ results in a decay of perturbation which is referred to as \enquote{good curvature}. Furthermore, such turbulent structures are called \textit{eddies} which have a typical radial scale of several Larmor radii. \cite{Newins2006}
The initial perturbation is growing if the normalized temperature gradient overcomes a critical value $R/L_\mathrm{T,c}$. For nonlinear turbulence with Cyclone Base Case parameters and circular concentric flux-surfaces the critical value is given by $R/L_\mathrm{T,c} \sim 6$. \cite{Dimits2000, Isliker2010}

### 2023-06-20 bachelorthesis/24-zonalflows.tex
#39: Bachelor thesis finished

\section{Zonal Flows and Shearing Rate $\wexb$}
\label{sec:zonalflow}

@@

\includegraphicsHere{Theory/Zonal-Flow-Generation.pdf}{
	Subdued turbulence because of stretched eddies, which breaks in smaller structures caused by sheared $\exb$ drifts \cite{Hammett2009}
}{fig:zonalflow}{0.9}

Furthermore, zonal flows are connected with the radially alternating zonal potential, which introduces a velocity shear, i.e., a radial variation of the zonal flow velocity. This causes a shear deformation of turbulent structures which is considered to play a significant role in the suppression of turbulence \cite{Biglari1990, Dimits2000}, especially the ion temperature gradient driven turbulence \cite{Nakata2012, Makwana2014, Maeyama2014, Whelan2018, Whelan2019}. The nonlinear generation mechanism causes a shift, also known as Dimits shift \cite{Dimits2000}, of the critical temperature gradient $R/L_\mathrm{T,c}$ which leads to a development of ITG driven turbulence at higher $\rlt$ compared to the linear threshold. The shearing process tilts and stretches turbulent eddies, caused by the shear flow, until they break into smaller eddies with reduced radial correlation length [Fig. \ref{fig:zonalflow}],  characterizing the shearing as a decorrelation process. Turbulence will then be reduced because of this process. \cite{Biglari1990,Diamond2005,Burnell1997} \bigskip

@@

The $\exb$ staircase pattern is manifest as radial structure formation in the $\exb$ shearing rate defined by\cite{Rath2016, Pueschel2008, Peeters2016}
\begin{equation}
	\wexb = \frac{1}{2} \frac{\partial^2 \langle \phi \rangle}{\partial \xcoord^2}~,
	\label{eq:shearingrate}
\end{equation}
where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature, $e$ is the elementary charge).
The zonal potential is calculated from the electrostatic potential $\phi$ on the two-dimensional $\xcoord$-$\ycoord$-plane at the low field side according to\cite{Rath2021}
\begin{equation}
\langle \phi \rangle = \frac{1}{L_\ycoord} \int_0^{L_\ycoord} \mathrm{d}\ycoord ~ \phi(\xcoord,\ycoord,s=0)~.
\end{equation}
The $\exb$ shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{Hahm1995, Waltz1998} and quantifies the zonal flow induced shearing of turbulent structures \cite{Biglari1990, Hahm1995, Burnell1997}. \\ \bigskip

As mentioned before the turbulent transport occurs through avalanches which also mediate the turbulence of the ITG driven turbulence in the adabatic electron approximation. Avalanches are initiated when the $\exb$ sharing rate has a zero crossing with a steep flank and propagates outwards with negative shearing and inwards with positive shearing. \cite{Idomura2009,McMillan2009} The fully developed staircase results in almost completely quenched turbulence. Its development results in a discontinuous step in the dependency of turbulent fluxes on the temperature gradient also known as \textit{finite heat flux threshold}. \cite{Peeters2016,Weikl2017}

>>>>

\newpage
\section{Zonal Flows and Shearing Rate $\wexb$}
\label{sec:zonalflow}

@@

\includegraphicsHere{Theory/Zonal-Flow-Generation.pdf}{
	Subdued turbulence because of stretched eddies, which breaks in smaller structures caused by sheared $\exb$ drifts. \cite{Hammett2009}
}{fig:zonalflow}{0.9}

Furthermore, zonal flows are connected with the radially alternating zonal potential, which introduces a velocity shear, i.e., a radial variation of the zonal flow velocity. This causes a shear deformation of turbulent structures which is considered to play a significant role in the suppression of turbulence \cite{Biglari1990, Dimits2000}, especially the ion temperature gradient driven turbulence \cite{Nakata2012, Makwana2014, Maeyama2014, Whelan2018, Whelan2019}. The nonlinear generation mechanism causes a shift, also known as Dimits shift \cite{Dimits2000}, of the critical temperature gradient $R/L_\mathrm{T,c}$ which leads to a development of ITG-driven turbulence at higher $\rlt$ compared to the linear threshold. The shearing process tilts and stretches turbulent eddies, caused by the shear flow, until they break into smaller eddies with reduced radial correlation length [Fig. \ref{fig:zonalflow}],  characterizing the shearing as a decorrelation process. Turbulence will then be reduced because of this process. \cite{Biglari1990,Diamond2005,Burnell1997} \bigskip

@@

The $\exb$ staircase pattern is manifest as radial structure formation in the $\exb$ shearing rate defined as\cite{Rath2016, Pueschel2008, Peeters2016}
\begin{equation}
	\wexb = \frac{1}{2} \frac{\partial^2 \langle \phi \rangle}{\partial \xcoord^2}~,
	\label{eq:shearingrate}
\end{equation}
where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature and $e$ is the elementary charge).
The zonal potential is calculated from the electrostatic potential $\phi$ on the two-dimensional $\xcoord$-$\ycoord$-plane at the low field side according to Ref. \citenum{Rath2021}
\begin{equation}
\langle \phi \rangle = \frac{1}{L_\ycoord} \int_0^{L_\ycoord} \mathrm{d}\ycoord ~ \phi(\xcoord,\ycoord,s=0)~.
\end{equation}
The $\exb$ shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{Hahm1995, Waltz1998} and quantifies the zonal flow induced shearing of turbulent structures \cite{Biglari1990, Hahm1995, Burnell1997}. \\ \bigskip

As mentioned before the turbulent transport occurs through avalanches which also mediate the turbulence of the ITG-driven turbulence in the adabatic electron approximation. Avalanches are initiated when the $\exb$ sharing rate has a zero crossing with a steep flank and propagates outwards with negative shearing and inwards with positive shearing. \cite{Idomura2009,McMillan2009} The fully developed staircase results in almost completely quenched turbulence. Its development results in a discontinuous step in the dependency of turbulent fluxes on the temperature gradient also known as \textit{finite heat flux threshold}. \cite{Peeters2016,Weikl2017}
